Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-23 paper/reviewer_response_PNAS.tex

6) The authors state that it there are more reports of escape from stalk antibodies for H3 than H1. This is simply not true. Actually, the first reference used as evidence for easy selection of H3 anti-stalk escape mutants (Chai et al) reported three escape mutants which a) did not disrupt binding completely and b) were crippled in vitro and in vivo. The authors should remove their statement as there is not systematic data to support this.

\response{We have removed this sentence from the paper.}

7) The biggest issue with the manuscript is the experimental set up. The authors titrate their rescued virus using a TCID50 assay on MDCK cells. Based on the titer that they measure, they passage the virus from the initial rescue once onto fresh MDCK cells using a low MOI. After infection, the virus dilution stays on the cells and is then harvested without centrifugation followed by RNA extraction and sequencing. There are many issue with this process.

\response{The reviewer has a number of questions and suggestions about the experimental design.
Some of these suggestions are things that were already implemented in our experiments and we simply failed to explain in adequate detail, while others are things that we have evidence are unimportant. 
Below we describe the issues one-by-one.

In addition, we have now included our \emph{full original lab notes} for the relevant experiments (\url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}) so that readers can examine everything that we did in full detail.
This level of transparency goes well beyond that usually taken in scientific papers, and will enable readers to fully understand our reasoning and procedures for each detail of the experiments.
}

@@

a) The TCID50 is likely underestimating the number of virus particles with HA genomic segments because not all of them (likely the minority) will be capable of inducting a full infectious cycle (but their presence might still be detected in the cell supernatants).

\response{The reviewer is certainly correct that TCID50 assays underestimate the number of HA genomic segments, as it is well known that the number of physical influenza virions exceeds the number of infectious units by 10-100 fold.
This is why we used an extremely low MOI (0.0035) so that most cells are still infected by no more than one virion.
It is impossible to grow a viral stock that is completely free of defective particles.
However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).

In short, the presence of non-infectious particles is an inherent feature of all virology work.
The best that can be done is to use MOIs and growth times that ensure that they do not dominate the experiments, and we have done that.
}

@@

b) The infectious inoculum is never washed off the infected cells meaning that any particles that are transferred from the rescue transfection are going to be detected during sequencing. This might even apply to naked RNPs released from dead cells.

\response{
We apologize for not expanding on this detail in the Methods of the original paper.

In fact, like the reviewer, we initially thought that a media change to wash aways the inoculum might be important. 
Therefore, in our experiments, we performed such a media change to remove the initial inoculum for biological replicates 1 and 2, and for one of the technical replicates (3-2) of biological replicate 3.
We did not perform such a change for the other technical replicate (3-2) of biological replicate 3.
It turns out that the media change makes no discernible difference, which is why we neglected to expand on this point in the original manuscript. 
We now have added the following explanatory text to the Methods:

\begin{quote}
\textsl{Three hours post-infection, we replaced the inoculum with fresh media for replicates 1, 2, and 3-2.
We did not perform a media change for replicate 3-1.
As can be seen in Figure~1D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
}
\end{quote}
The full details of the media changes can be found in the lab notes we have included at We have also included in the GitHub repository a PDF of the lab notebook describing the rescue and passage of the mutant virus libraries at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25).

We think that the reason that washing away the original inoculum has no discernible effect on the results it that the low concentration of virus in the inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing, since vastly more new virions are produced from the infected cells.
}

@@

c) This would actually not be an issue, if the dilution from the rescue culture to the new culture would be large enough. But this is not the case. The authors passaged $9\times10^5$ TCID50 (and likely much more DIs/RNA) from the rescue cultures onto the new cells in a volume of 25 ml. This is, depending on experiment, between 10 and 1 ml of supernatant of the original rescue and leads to a 1:2.5 or 1:25 dilution respectively.

\response{
We apologize if the original wording of our methods were unclear. 
For each mutant virus library, we passaged a total of $9\times10^5$ TCID50, but this was passaged across \emph{fifteen} 15-cm dishes \emph{each} of which received 25 ml.
Because we used an inoculum of 2.5 TCID50 / $\mu$l across these fifteen dishes, this corresponds to dilutions that range from 1:37 to 1:294.
The number of dishes used to passage each mutant virus library is now mentioned in the Methods; before it had to be calculated from the stated total number of cells and the cells per dish.

Also, we note that the supernatants from the helper virus rescues are collected at just 24 hours post-infection, which leads to lower viral titers (which is why we have to passage relatively large volumes), but also presumably much lower amounts of DIs / RNA.
This is probably the reason why (as described above), we did removing the inoculum did not yield discernibly affect the results.
}

@@

d) The supernatant is never centrifuged before RNA extraction. Cell debris is not removed which is likely contributing significantly to RNA sequences detected in the supernatant. However, RNA associated with cell debris is not representative of RNA from fit, replicating viruses.

\response{We apologize for failing to provide sufficiently detailed methods stating that we ultra-centrifuged \emph{clarified} the viral supernatant before RNA extraction.
The supernatant was clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris, similar to the reviewer's suggestions.
We have updated the Methods to now include this detail, and (as mentioned above) also now include our full lab notes so that all other details can be reviewed.
Specifically, see \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25)
}

@@

The experiments should be repeated with at least one more passaging step, proper removal of inoculum and washing of the cells after infection. Centrifugation steps should also be included between the passages and before RNA extraction in the end.

\response{For the reasons described above, it is not necessary to totally repeat the experiments.
We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
\begin{itemize}
\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
\end{itemize}
For any experimental study there are an almost infinite number of slight methodological variations that could be employed, and it is impossible to test them all.
Our experiments used reasonable parameter choices, were performed in full biological triplicate, and give meaningful / useful results with respect to the larger goals of the study.}

### 2018-06-23 paper/manuscript.tex
respond to reviewer 1 comment 7

We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
We collected viral supernatant for sequencing 48 hours post-infection.

@@

>>>>

We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
We did not perform a media change for replicate 3-1.
As can be seen in Figure~\ref{fig:dms_overview}D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
We collected viral supernatant for sequencing 48 hours post-infection.

Full lab notes from these experiments are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.

@@

Full lab notes for the library preparation are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.
